---
title: "Loading Expedition Data"
author: "A. Warmack"
date: "9/4/2018"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

#Summary
The boat performance data was collected using 

# Combining Log Files
```{r}
dat_folder <- "../raw_data/expedition_data"

#get the list of log files
dat_files <- list.files(dat_folder, full.names = TRUE)

#Read the csv files
exp_raw <- lapply(dat_files, read.csv, stringsAsFactors=FALSE)
exp_raw <- do.call(rbind, exp_raw)
```



##Cleaning Up The Data
###Removing Unneeded Columns
The Expedition data contains many columns that do not contain any data. 
```{r}
#Find columns where all is NA
allna <- which(apply(exp_raw, 2, function(x) all(is.na(x))))   

#remove those columns
expdat <- exp_raw %>% select(-allna)
```



###Convert the time from Excel Format
```{r}
expdat$time <- as.POSIXct((expdat$Utc) * (60*60*24), origin = "1899-12-30", tz = "America/Detroit")
```

###Select the useful columns
- *time*
- *Bsp* = Boatspeed thru water
- *Awa* = Apparent Wind Angle (Degrees Relative to boat)
- *Aws* = Apparent Wind Speed (in knots)
- *Twa* = True Wind Angle (Degrees Relative to boat)
- *Tws* = True Wind Speed (in knots)
- *Twd* = True Wind Direction (Bearing Relative to the Earth)
- *Lat/Lon* = Position
- *Cog* = Course Over Ground (Bearing in Degrees)
- *Sog* = Speed Over Ground (in Knots)
```{r, echo=TRUE}
#Select only needed columns
expdat <- expdat %>% select(time, Bsp, Awa, Aws, Twa, Tws, Twd, Lat, Lon, Cog, Sog)

#change names to lowercase for easier handling
names(expdat) <- tolower(names(expdat))

str(expdat)
```

###Remove Data from before and after the race
```{r, echo=TRUE}
#remove pre-start and post-finsih
expdat <- expdat %>% dplyr::filter(time > as.POSIXct("2018-7-14 16:00:00"))

expdat <- expdat %>% dplyr::filter(time < as.POSIXct("2018-07-17"))

head(expdat$time)
```

##Data Exploration
### Basic Stats
```{r}
summary(expdat)
```

We can see there is a large number of NA's for most columns. The most important columns are apparent wind angle (awa), apparent wind speed (awa), boatspeed (bsp), and position (lat/lon). The remaining variables can be calculated from these values.  

#where do we have AWS and AWA
```{r}
library(ggplot2)

expdat$index <- 1:nrow(expdat)

ggplot(expdat) + geom_point( aes(x=index, y=awa))
ggplot(expdat) + geom_point(aes(x=index, y=aws))
```


##Calcuale TWD for missing variables
```{r}
fromNorth <- function(bearing){
  bearing <- normbear(bearing)  #first convert to a normal bearing
  
  fromNorth <- ifelse(bearing > 180, bearing-360, bearing)
  fromNorth <- ifelse(fromNorth < -180, 360+fromNorth, fromNorth)
  return(fromNorth)
}

normbear <- function(x){
  #converts back to a normal bearing
  x %% 360
}

getTWD <- function(twa, hdg){
  #twa = TRUE WIND ANGLE
  #hdg = Heading
  hdg <- fromNorth(hdg)
  twa <- fromNorth(twa)
  
  normbear(hdg + twa)
  
}

missingtwd <- is.na(expdat$twd)

expdat$twd[missingtwd] <- getTWD(expdat$twa[missingtwd] , hdg = expdat$sog[missingtwd] )

```

Handle Outlier Data
```{r}

expdat$twa[expdat$twa > 180] <- NA
expdat$tws[expdat$tws > 25]  <- NA
expdat$bsp[expdat$bsp > 15]  <- NA
expdat$sog[expdat$sog > 15]  <- NA

```


##Calculate the performance to our polar
The polar provides the optimal speed given an twa and tws.
```{r}
library(akima)

load("../raw_data/polarmodel.rda")

#get optimal boat speed at a given TWA (=btw)
getOptV <- function(btw, vtw, pol.model){
  
  #returns Optimal Velocity at a given Bearing to Wind (BTW) and Wind Speed (VTW)
  btw <- normbear(btw)
  
  btw <- fromNorth(btw)

  btw <- abs(btw)
  
  #output the optimum Velocity for a given BTW and wind speed. 
  v <- bilinear(x=pol.model$twa, y=pol.model$tws, z=pol.model$pol.matrix, x0=btw, y0=vtw)
  
  return(v$z)
  
}

#note that this will only work if all of the input data is an actual number
hasdata <- !is.na(expdat$twa) & !is.na(expdat$tws) & abs(expdat$twa)>40

expdat$opt_v[hasdata] <- getOptV(btw=expdat$twa[hasdata], vtw = expdat$tws[hasdata], pol.model = pol.model)

#Calculate the percentage of our polar
expdat$pol_perc <- (expdat$bsp / expdat$opt_v) * 100



```

```{r}
expdat <- na.omit(expdat)

expdat <- expdat[abs(expdat$twa)!=180, ]
```


```{r}
#histogram of our performance
ggplot(expdat) + geom_histogram(aes(x=expdat$pol_perc), bins=50)

```

```{r}
ggplot(expdat) + 
  geom_point(aes(x=twa, y=pol_perc)) + 
  scale_x_continuous(limits=c(-180, 180))+
  scale_y_continuous(limits=c(100,250))
```

```{r}
ggplot(expdat) + 
  geom_point(aes(x=tws, y=pol_perc), alpha=0.1) + 
  scale_x_continuous(limits=c(0, 20)) + 
  scale_y_continuous(limits=c(0, 500)) + 
  geom_hline(aes(yintercept=100), color="green")

```


```{r}

ggplot(expdat) + geom_point(aes(x=tws, y=bsp), alpha=0.1) + geom_point(data=pol, aes(x=tws, y=opt_v, group=tws), color="red" )

```

```{r}
ggplot(expdat) + 
  geom_point(aes(x=twa, y=bsp, color=pol_perc), alpha=0.1)+
  geom_path(data=pol, aes(x=twa, y=opt_v, group=tws), color="red" ) +
  scale_color_gradient2(midpoint = 100, mid="green", low="yellow", high="blue", limits=c(50,150))+
  coord_polar(start = -pi)
```


```{r}
expdat$twa_cut <- cut_number(abs(expdat$twa), 4)
expdat$tws_cut <- cut_number(expdat$tws, 4)

ggplot(expdat) + 
  geom_histogram(aes(x=pol_perc), binwidth=2)+
  scale_x_continuous(limits=c(0,150))+
  facet_grid(tws_cut ~ twa_cut)
```

#Calculating Optimum VMC

## Calculate bearing to the mark

### Load Marks
```{r}
marks <- read.csv("../raw_data/marks.csv", stringsAsFactors = FALSE)
marks
```

```{r}
whichMark <- function(lat, marks, returns="mark"){
  
  if(lat > max(marks$lat)){
    #If the lat is greater than the maximum latitude of all marks, return the last mark
    return(marks[nrow(marks), returns])
  }
  
  #Else return the first mark that has a lat greater than the current position
  mark <- marks[min(which(lat < marks$lat)), returns]
  return(mark)
}

expdat$mark <- sapply(expdat$lat, whichMark, marks, returns="mark")
expdat$markLat <- as.numeric(sapply(expdat$lat, whichMark, marks, returns="lat"))
expdat$markLon <- as.numeric(sapply(expdat$lat, whichMark, marks, returns="lon"))
```


```{r}
#get states shape files
mi <- map_data("state") %>% filter(region=="michigan")

mimap <- ggplot()+
          geom_polygon(data=mi, aes(long, lat, group=group)) + 
          coord_map("mercator", xlim=c(-85, -82), ylim=c(42.8, 46))

mimap + geom_point(data=expdat, aes(x = lon, y=lat, color=mark))

```

#get the bearing to the mark
```{r}
library(geosphere)

##Calculate Bearing to the Mark
p1 <- expdat[, c("lon", "lat")]
p2 <- expdat[, c("markLon", "markLat")]
  
expdat$btm <- bearingRhumb(p1, p2)

```


#Calculating Optimum VMC
```{r}
getVMC <- function(btm, cog, bsp){
  #given a bearing to the mark (BTM) and course over ground(COG)
  # what is the vmc being made towards the mark
  
  #get angle between btm and cog
  angle <- fromNorth(cog) - fromNorth(btm)
  
  vmc <- bsp * cos(angle*(pi/180))
  
  return(vmc)
  
}

getTWA <- function(btm, twd){
  #twd <- fromNorth(twd)
  #btm <- fromNorth(btm)
  
  fromNorth(twd-btm)
  
}

optvmc <- function(btm, twd, tws, pol.model){
  #BTM = Bearing to Mark (0-360)
  #TWD = True Wind Direction (0-360)
  #
  
  col_names <-  c("btm", "twd", "tws", "opt_twa", "opt_bear", "degoffmark", "opt_bsp", "opt_vmc", "vmc_gain")
  
  if(any(is.na(c(btm, twd, tws)))){ 
    
    na_row <- as.data.frame(matrix(NA, nrow=1, ncol=9))
    names(na_row) <- col_names
    
    return(na_row)
  }
  
  
  btm <- normbear(btm)
  twd <- normbear(twd)
  
  #nominal TWA (pointing at mark) where 0 at the wind
  ntwa <- getTWA(btm, twd)
  
  #create vector of angles off the mark (0 at the wind)
  off_mark <- seq(-50,50, by=1)         #degrees away from the mark + = Stbd
  twa_s <- fromNorth(ntwa + off_mark)   #twa at each degree off the mark
  
  #bearings of each test position (actual direction of optimal VMC)
  opt_bear <- normbear(btm - off_mark)
  
  tws <- rep(tws, length(twa_s))
  
  #get BSP at each twa
  bsp <- getOptV(twa_s, vtw = tws, pol.model)
  
  #get vmc at each angle off mark
  vmc <- bsp * cos(off_mark*(pi/180))
  
  #find max VMC
  vmcopt <- max(vmc)
  
  #find gain
  if(abs(ntwa) < 40){  ntwa <- 40 }   #if we're on a beat, assume close hulled
  
  vmc_gain <- vmcopt - getOptV(ntwa, vtw=tws[1], pol.model)
  
  opt <- data.frame(btm, twd, tws, twa_s, opt_bear, off_mark, bsp, vmc, vmc_gain)
  names(opt) <- col_names
  
  return_opt <- head(opt[which(vmcopt==vmc), ], 1)   #only return one value if there are more than 1
  
  return(return_opt)
  
}

optimalcourse <- mapply(FUN = optvmc, expdat$btm, expdat$twd, expdat$tws, MoreArgs = list(pol.model), SIMPLIFY = FALSE)

optimalcourse <- do.call(bind_rows, optimalcourse)
  
expdat <- bind_cols(expdat, optimalcourse[, 4:9])

```

#Summary of VMC vs. path sailed
```{r}

expdat$vmc <- getVMC(expdat$btm, cog=expdat$cog, bsp = expdat$bsp)

expdat$off_vmc <- expdat$vmc - expdat$opt_vmc

ggplot(expdat) + geom_histogram(aes(x=off_vmc))+scale_x_continuous(name="Actual VMC Vs. Optimal VMC [kts]")

```

```{r}
expdat$degoffvmc <- expdat$twa - expdat$opt_twa

ggplot(expdat) + geom_histogram(aes(x=degoffvmc), binwidth=5)+scale_x_continuous(name="Degrees from Optimal VMC angle")

```

```{r}
mimap + geom_point(data=expdat, aes(x=lon, y=lat, color=degoffvmc)) + scale_color_gradient2(mid="yellow")
```

```{r}
mimap + geom_point(data=expdat[expdat$mark=="S3 TBLT", ], aes(x=lon, y=lat, color=twd)) + scale_color_gradient2(mid="yellow") + coord_map("mercator", xlim=c(-84, -82), ylim=c(44, 45))

```

#write the full data
```{r}
write.csv(expdat, "../mod_data/expdition_full.csv", row.names = FALSE)
```


#We need to summarize the data for printing. 
```{r}
library(lubridate)

exp_5m <- expdat %>% 
            mutate(time = round_date(time, unit = "5 mins")) %>% 
            mutate(cog = fromNorth(cog)) %>%   #should average as degrees from north
            group_by(time) %>% 
            summarise_if(is.numeric, mean, na.rm=TRUE) %>%
            mutate(cog = normbear(cog))

exp_15m <- expdat %>% 
            mutate(time = round_date(time, unit = "15 mins")) %>% 
            mutate(cog = fromNorth(cog)) %>%   #should average as degrees from north
            group_by(time) %>% 
            summarise_if(is.numeric, mean, na.rm=TRUE) %>%
            mutate(cog = normbear(cog))

ggplot() + geom_point(data=expdat,aes(x=time, y=cog), color="blue" ) + geom_point(data=exp_5m, aes(x=time, y=cog))  

#mimap + geom_point(data=exp_5m, aes(x=lon, y=lat))

write.csv(exp_5m, "../mod_data/expdition_5minavg.csv", row.names = FALSE)
write.csv(exp_15m, "../mod_data/expdition_15minavg.csv", row.names = FALSE)

```

```{r}
exp_5m$twa_cut <- cut_number(abs(exp_5m$twa), 4)
exp_5m$tws_cut <- cut_number(exp_5m$tws, 4)

ggplot(exp_5m, aes(x=twa, y=bsp, color=tws_cut)) + 
  geom_point()  + coord_polar(start = -pi) + facet_wrap(. ~ tws_cut)
```




